\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-08 chapters/stbf_struct.tex
checking bttda
Proofread

  \emph{This chapter, except section~\ref{sec:stbf-struct-kronlda}, was
  published as~\textcite{VanDenKerchove2022}}.

@@

	%Brain-computer interfaces (BCIs) establish a direct communication pathway
	%between the brain and an external device~\cite{Wolpaw2002}.
	%Severely disabled patients with impaired or absent communication capabilities
	%can benefit from BCIs to restore normal functioning~\cite{Naci2012,Chaudhary2016}.
	%BCIs can be implemented in multiple ways, using non-invasive recording
	%techniques such as electroencephalography (EEG)~\cite{Abiri2019},
	%magnetoencephalography (MEG)~\cite{Mellinger2007},
	%functional Near-Infrared Spectroscopy (fNIRS)~\cite{Hong2015}, and Optically Pumped
	%Magnetometers (OPM MEG)~\cite{Paek2020}, or semi-invasive and invasive methods such as
	%electrocorticography (ECoG)~\cite{Schalk2011} or microelectrode
	%arrays~\cite{Maynard1997} which require surgery to implant a recording device.
	%While invasive BCIs yield the highest information transfer
	%rate~\cite{Willett2021}, non-invasive BCIs are preferable for short-term use
	%since they are not susceptible to the risks that come with surgery.
	%Of the non-invasive options, EEG is the most cost-effective and practical as it
	%is not limited to the same controlled settings as MEG and OPM MEG.

	%Besides the recording method, BCIs differ in the communication paradigms used
	%for communication~\cite{Abiri2019}.
	%A popular class of BCI paradigms relies on the evocation
	%of Event-Related Potentials (ERPs) in the brain in response to visual, auditory, or tactile stimulation, given their low decoding cost and generally short
	%calibration time before usage~\cite{Gao2014, Kapgate2015}.
	%The study we report on focuses on the visual P300 oddball ERP in response to a
	%rare but attended visual stimulus.
	%The decoder detects whether this ERP is present to determine which stimulus
	%the user attended.
	%The P300 paradigm has been used extensively in BCI development and is easy to
	%set up~\cite{Farwell1988, Sellers2006, Barachant2014, Philip2020}.

	There are multiple state-of-the-art P300 classification methods, like Support
	Vector Machines (SVMs)~\cite{Tayeb2014}, deep
	learning models~\cite{Vareka2020,Borra2020}, and Riemannian Geometry
	classifiers~\cite{Barachant2014}.
	While these models often return a high classification accuracy, there is a need
	for lightweight models -- lightweight models lead to fast off-line analyses and can be transferred to consumer-grade hardware.
	When moving towards plug-and-play solutions, BCI calibration sessions should be short and model training times low.
	The spatiotemporal beamformer~\cite{VanVliet2015, Wittevrongel2016} belongs to
	this class of ERP decoding models as it achieves state-of-the-art performance and is fast to train.
	Earlier work has shown that it is
	possible to apply the spatiotemporal beamformer to multiple time-locked visual
	BCI paradigms, including the P300 oddball paradigm,
	Steady-State Visually Evoked Potentials (SSVEP), code-modulated Visually Evoked Potentials (cVEP)~\cite{Wittevrongel2017a} and
	motion-onset Visually Evoked Potentials (mVEP)~\cite{Libert2021}.

	This work shows that the original spatiotemporal
	beamformer~\cite{Wittevrongel2016} can fall short in performance when BCI
	calibration data are restricted.
	We also show that the spatiotemporal beamformer does not scale well for
	higher spatial and temporal resolution cases.
	As a response to these issues, we introduce a regularization method that
	exploits prior knowledge about the spatiotemporal nature of the EEG signal to
	improve the accuracy for low data availability settings and speed up the
	classifier training time, thereby considerably reducing memory usage.
	Similar structured regularization approaches have been applied to other linear
	ERP classifiers~\cite{GonzalezNavarro2017, Vliet2020} and have shown
	significant increases in performance.
	Additionally, we show that regularization results in an interpretable
	classification model, which can aid in analyzing and developing spatiotemporal beamformer-based classifiers.

	\section{Materials \& methods}
	\subsection{Notation}
	We represent matrices with cursive capital letters, vectors with bold
	lowercase letters, and scalars with cursive lowercase letters.
	The epoched EEG data with $n$ epochs, $c$ channels, and $s$ samples are
	represented in epoch format as $\{X_i\in\mathbb{R}^{c\times s}\}^n_{i=1}$ or flattened vector format by concatenating all channels for each epoch.
	Flattening results in $\{\mathbf{x}_i\in\mathbb{R}^{cs}\}^n_{i=1}$ such that $\mathbf{x}_i = \text{vec}(X_i)$.
	The real covariance matrix of the epochs in vector format is
	denoted by $C$, estimators thereof as $\hat{C}$.

	\subsection{Spatiotemporal beamforming}
	LCMV-beamforming was initially introduced to EEG analysis as a filter for
	source localization~\cite{VanVeen1997} to enhance the signal-to-noise ratio
	(SNR).
	Van Vliet et al.~\cite{VanVliet2015} first applied the spatiotemporal
	LCMV-beamformer as a method for the analysis of ERPs.
	The extension to the combined spatiotemporal domain~\cite{VanVliet2015} and the
	data-driven approaches proposed by Treder et al.~\cite{Treder2016} and
	Wittevrongel et al.~\cite{Wittevrongel2016} allow for its application to classification problems.

	For the following analyses, we assume that all EEG channels are normalized with zero mean and unit variance without loss of generality.
	Solving \autoref{eq:minimum_variance} under the linear constraint given by
	\autoref{eq:linear_constraint} returns the filter weights $w$ defining the spatiotemporal LCMV-beamformer.
	\begin{equation}
		\underset{\mathbf{w}}{\arg\min}\mathbf{w}^\intercal C
		\mathbf{w}^\intercal
		\label{eq:minimum_variance}
	\end{equation}
	\begin{equation}
		\mathbf{a}^\intercal\mathbf{w} = 1
		\label{eq:linear_constraint}
	\end{equation}

	These weights minimize the variance of the output of the filter while enhancing
	the signal characterized by the constraint.
	$\mathbf{a} = \text{vec}(A)$ is the data-driven activation pattern, a template
	of the signal of interest maximizing the difference between two classes of
	epochs, determined as follows:

@@

	The method of Lagrange multipliers then gives the closed-form solution to the minimization problem posed by
	\autoref{eq:minimum_variance} and \autoref{eq:linear_constraint} as:

	\begin{equation}
		\mathbf{w} =
		\frac{C^{-1}\mathbf{a}^\intercal}
		{\mathbf{a}C^{-1}\mathbf{a}^\intercal}
		\label{eq:closed_form}
	\end{equation}

@@

	resulting in a scalar output per epoch.
	The linear constraint in \autoref{eq:linear_constraint} ensures that the
	beamformer maps epochs containing a target response to a score close to one
	and, conversely, epochs not containing a target response to a score close to
	zero.

	\subsection{Covariance matrix regularization}
	While the spatiotemporal beamformer, in theory, achieves optimal separation
	between target and non-target classes, in analogy to linear discriminant analysis~\cite{TREDER2016279}, it does not always perform well on unseen data.
	The main challenge is to find a good estimator for the inverse covariance matrix $C^{-1}$ since the real underlying covariance matrix generating the data is, in principle, unknown.

@@

	\begin{equation}
		\hat{C}_\text{emp} =
		\frac{1}{n-1}\sum^{n}_{i=1}\mathbf{x}_i\mathbf{x}_i^\intercal
		\label{eq:empirical_covariance}
	\end{equation}

	\begin{equation}
		\widehat{C^{-1}}_\text{emp} = \hat{C}_\text{emp}^+
		\label{eq:empirical_inverse_covariance}
	\end{equation}

	The Moore-Penrose pseudoinverse $^+$, ensures a solution exists when $\hat{C}_\text{emp}$ is singular.
	\autoref{fig:kronlda-covs}a and \autoref{fig:kronlda-covs}b respectively show examples of the
	empirical estimators of the covariance and the inverse covariance matrices.
	The empirical estimator suffers from performance and
	stability issues if the number of epochs $n$ used or estimation is not much larger than the number of features $cs$~\cite{Stein1956,Khatri1987}.

@@

	\begin{equation}
		\hat{C}_\alpha =
		(1-\alpha) \hat{C}_\text{emp}
		+ \alpha\frac{\text{Tr}(\hat{C}_\text{emp})}{cs}\mathbb{I}
		\label{eq:shrinkage}
	\end{equation}

	\begin{equation}
		\widehat{C^{-1}}_\alpha =
		\hat{C}^+_\alpha
		\label{eq:shrinkage_inverse}
	\end{equation}

	with $0<\alpha<1$.
	Analogous to L2 regularization of the beamforming problem,
	shrinkage reduces the ratio between the smallest and largest eigenvalues
	of the covariance matrix by strengthening the diagonal.
	\autoref{fig:kronlda-covs}c and \autoref{fig:kronlda-covs}d respectively show examples of the
	shrunk estimator of the covariance and the inverse covariance matrices.

	Earlier work~\cite{Libert2021} applied shrinkage regularization to ERP
	decoding with the spatiotemporal beamformer and showed competitive performance
	compared to other state-of-the-art decoding techniques like stepwise LDA or SVM.
	The abovementioned work chooses the shrinkage coefficient $\alpha$ as a fixed hyperparameter.
	However, its optimal value depends on the number of training epochs, the
	covariance matrix's dimensionality, and the independence and variance of the
	data, which can vary across evaluation settings and per session.
	The optimal value for $\alpha$ can be found with a line search using cross-validation, but this can be a costly procedure.
	Methods exist to estimate an optimal shrinkage value from the data directly.
	Most notable among these are the Ledoit-Wolf procedure~\cite{Ledoit2004},
	Rao-Blackwell Ledoit-Wolf~\cite{Chen2010shrinkage}, and Oracle Approximating Shrinkage~\cite{Chen2010shrinkage}.
	A more recent estimation method~\cite{Tong2018} emulates a leave-one-out
	cross-validation (LOOCV) scheme expressed by the data-driven closed-form
	estimate:

	\begin{equation}
		\alpha =
		1-\frac{
			\frac{n}{n-1}\text{Tr}(\hat{C}_\text{emp}^2)
			- \frac{2}{cs}\left[\text{Tr}(\hat{C}_\text{emp})\right]^2
			+ \frac{1}{cs}\text{Tr}(\hat{C}_\text{emp}^2)
			- \frac{1}{n(n-1)}\sum_{i=1}^n||\mathbf{x}_i||_2^4
		}
		{
			\frac{n^2 -2n}{(n-1)^2}\text{Tr}(\hat{C}_\text{emp}^2)
			- \frac{2}{cs}\left[\text{Tr}(\hat{C}_\text{emp})\right]^2
			+ \frac{1}{cs}\text{Tr}(\hat{C}_\text{emp}^2)
			+ \frac{1}{n(n-1)^2}\sum_{i=1}^n||\mathbf{x}_i||_2^4
		}
		\label{eq:loocv}
	\end{equation}

	We opt for the LOOCV shrinkage estimator because it avoids some of the
	assumptions made by~\cite{Ledoit2004} and~\cite{Chen2010} and
	because it generalizes to structured covariance estimation as described in
	\autoref{seq:structured_estimation}.

	\subsubsection{Spatiotemporal beamforming with Kronecker-Toeplitz structured covariance}
	\label{seq:structured_estimation}
	Exploiting prior knowledge about the spatiotemporal structure of the EEG signal leads to a more regularized estimator of the covariance.
	When viewing the example of empirical spatiotemporal EEG covariance in
	\autoref{fig:kronlda-covs}a, it becomes clear that this matrix consists of a block pattern of repeated, similar matrices.
	Due to the multi-channel nature of the signal, we assume that the covariance of spatiotemporal EEG epochs is a Kronecker
	product of two smaller
	matrices~\cite{Munck1992,DeMunck1999,Huizenga2002}, as expressed
	by:

	\begin{equation}
		\hat{C}_\text{struct} = \hat{S} \otimes \hat{T}
		\label{eq:kronecker}
	\end{equation}

	with $\otimes$ the Kronecker product operator.
	$\hat{S} \in \mathbb{R}^{c\times c}$ and $\hat{T} \in \mathbb{R}^{s\times s}$ respectively correspond to estimators of the spatial and temporal covariance of the data.
	Furthermore, because the temporal covariance of the EEG-signal is
	stationary (i.e., it is only dependent on interval length between covarying
	time samples)~\cite{Bijma2003}, it is assumed to have a Toeplitz-matrix structure:

	\begin{equation}
		\hat{T}_{i,j} = \hat{T}_{i+1,j+1}
		\label{eq:toeplitz}
	\end{equation}

	\autoref{prop:inverse_kronecker} then leads to
	\autoref{eq:cov_inverse_kronecker} to estimate the inverse
	covariance.
	\begin{property}
		$(U \otimes V)^+ = U^+ \otimes V^+$ for any non-singular
		matrices $U$ and $V$~\cite{Langville2004}.
		\label{prop:inverse_kronecker}
	\end{property}

	\begin{equation}
		\widehat{C^{-1}}_\text{struct} = \hat{S}^+ \otimes \hat{T}^+
		\label{eq:cov_inverse_kronecker}
	\end{equation}

	Finally, based on \autoref{prop:kronecker_multiplication},
	\autoref{eq:closed_form} can be reformulated more efficiently as
	\autoref{eq:closed_form_kronecker}.
	\begin{property}
		$(U\otimes V)\cdot\text{vec}(W) = \text{vec}(VWU^\intercal)$
		for any matrices $U\in\mathbb{R}^{p\times p}$,
		$V\in\mathbb{R}^{q\times q}$ and $W\in\mathbb{R}^{p\times q}$~\cite{Loan2000}.
		\label{prop:kronecker_multiplication}
	\end{property}

	\begin{equation}
		\hat{\mathbf{w}}_\text{struct} =
		\frac{\hat{S}^+A^T\hat{T}^+}
		{\mathbf{a}\cdot\text{vec}(\hat{S}^+A^T\hat{T}^+)}
		\label{eq:closed_form_kronecker}
	\end{equation}

	Using \autoref{eq:closed_form_kronecker} removes the need to calculate the full, high dimensional Kronecker product $\hat{S}^+\otimes
		\hat{T}^+$.
	\autoref{fig:kronlda-covs}e and \autoref{fig:covs}f respectively show examples of the
	structured covariance and inverse covariance estimators,
	consisting of a spatial Kronecker factor (\autoref{fig:kronlda-covs}g and
	\autoref{fig:kronlda-covs}h) and a temporal component (\autoref{fig:kronlda-covs}i and
	\autoref{fig:kronlda-covs}j).

	\begin{figure}
		\includegraphics[width=\linewidth]{figures/stbf_struct/covs.eps}
    \caption[Estimated covariance and inverse covariance.]{Different estimators of the covariance and inverse covariance
			of 100 epochs of data from \textit{Subject 01} for channels
			\textit{Fz}, \textit{Cz}, \textit{Pz}, and \textit{Oz} and time samples between 0.1s and 0.6s.
			Regularized estimators of the inverse covariance exhibit less extreme values and have a sparser structure.
			(\textbf{a,b}) Empirical covariance and inverse covariance matrices.
			(\textbf{c,d}) Shrunk covariance and inverse covariance matrices with $\alpha=0.14$ as
			determined by the closed-form LOOCV method. (\textbf{e,f}) Kronecker-Toeplitz
			structured covariance and inverse covariance matrices.
			(\textbf{g,h}) Spatial Kronecker factor of the Kronecker-Toeplitz structured shrunk estimator and its inverse.
			(\textbf{i,j}) Temporal Kronecker factor of the Kronecker-Toeplitz structured shrunk estimator and its inverse.}
		\label{fig:kronlda-covs}
	\end{figure}

	The Kronecker approach has shown significant performance yields in different linear spatiotemporal EEG and MEG
	applications~\cite{DeMunck2002,Huizenga2002,Beltrachini2013,GonzalezNavarro2016,GonzalezNavarro2017}.
	Van Vliet \& Salmelin~\cite{Vliet2020} have applied a Kronecker-structured covariance estimator to ERP classification with linear models in a post-hoc fashion.
	Our work goes further by embedding the Kronecker structure in the
	spatiotemporal beamformer training process, using a data-adaptive shrinkage
	method, and regularizing the covariance further by imposing a Toeplitz
	structure on the temporal covariance.

@@

	\begin{subequations}
		\begin{equation}
			\tilde{S}_{k+1} =
			\frac{1}{n}
			\sum^n_{i=1}X_i^\intercal\hat{T}_k^+X_i
			\label{eq:fpi_spatial}
		\end{equation}
		\begin{equation}
			\tilde{T}_{k+1} =
			\frac{1}{n}
			\sum^n_{i=1}X_i\hat{S}_k^+X_i^\intercal
			\label{eq:fpi_temporal}
		\end{equation}
	\end{subequations}

	\begin{subequations}
		\begin{equation}
			\tilde{S}_{k+1}^{(\beta)} =
			(1-\beta_{k+1})\tilde{S}_{k+1}
			+\beta_{k+1}\frac{\text{Tr}(\tilde{S}_{k+1})}{c}\mathbb{I}
			\label{eq:fpi_spatial_shrunk}
		\end{equation}
		\begin{equation}
			\tilde{T}_{k+1}^{(\gamma)} =
			(1-\gamma_{k+1})\tilde{T}_{k+1}
			+\gamma_{k+1}\frac{\text{Tr}(\tilde{T}_{k+1})}{s}\mathbb{I}
			\label{eq:fpi_temporal_shrunk}
		\end{equation}
	\end{subequations}

>>>>

  \emph{This chapter was
  published as~\textcite{VanDenKerchove2022}}.

@@

	%Brain-computer interfaces (\acp{bci}) establish a direct communication pathway
	%between the brain and an external device~\cite{Wolpaw2002}.
	%Severely disabled patients with impaired or absent communication capabilities
	%can benefit from \acp{bci} to restore normal functioning~\cite{Naci2012,Chaudhary2016}.
	%\acp{bci} can be implemented in multiple ways, using non-invasive recording
	%techniques such as electroencephalography (\ac{eeg})~\cite{Abiri2019},
	%magnetoencephalography (MEG)~\cite{Mellinger2007},
	%functional Near-Infrared Spectroscopy (fNIRS)~\cite{Hong2015}, and Optically Pumped
	%Magnetometers (OPM MEG)~\cite{Paek2020}, or semi-invasive and invasive methods such as
	%electrocorticography (ECoG)~\cite{Schalk2011} or microelectrode
	%arrays~\cite{Maynard1997} which require surgery to implant a recording device.
	%While invasive \acp{bci} yield the highest information transfer
	%rate~\cite{Willett2021}, non-invasive \acp{bci} are preferable for short-term use
	%since they are not susceptible to the risks that come with surgery.
	%Of the non-invasive options, \ac{eeg} is the most cost-effective and practical as it
	%is not limited to the same controlled settings as MEG and OPM MEG.

	%Besides the recording method, \acp{bci} differ in the communication paradigms used
	%for communication~\cite{Abiri2019}.
	%A popular class of \ac{bci} paradigms relies on the evocation
	%of Event-Related Potentials (\acp{erp}) in the brain in response to visual, auditory, or tactile stimulation, given their low decoding cost and generally short
	%calibration time before usage~\cite{Gao2014, Kapgate2015}.
	%The study we report on focuses on the visual P300 oddball \ac{erp} in response to a
	%rare but attended visual stimulus.
	%The decoder detects whether this \ac{erp} is present to determine which stimulus
	%the user attended.
	%The P300 paradigm has been used extensively in \ac{bci} development and is easy to
	%set up~\cite{Farwell1988, Sellers2006, Barachant2014, Philip2020}.

  There are multiple state-of-the-art P300 classification methods, like
  \acp{svm}~\cite{Tayeb2014}, deep
	learning models~\cite{Vareka2020,Borra2020}, and Riemannian Geometry
	classifiers~\cite{Barachant2014}.
	While these models often return a high classification accuracy, there is a need
	for lightweight models -- lightweight models lead to fast off-line analyses and can be transferred to consumer-grade hardware.
	When moving towards plug-and-play solutions, \ac{bci} calibration sessions should be short and model training times low.
  The \ac{stbf}~\cite{VanVliet2015, Wittevrongel2016} belongs to
	this class of \ac{erp} decoding models as it achieves state-of-the-art performance and is fast to train.
	Earlier work has shown that it is
	possible to apply the spatiotemporal beamformer to multiple time-locked visual
  \ac{bci} paradigms, including the P300 oddball paradigm, \ac{ssvep},
  \ac{cvep}~\cite{Wittevrongel2017a}, \ac{mvep}~\cite{Libert2021}.

	This work shows that the original spatiotemporal
	beamformer~\cite{Wittevrongel2016} can fall short in performance when \ac{bci}
	calibration data are restricted.
	We also show that the spatiotemporal beamformer does not scale well for
	higher spatial and temporal resolution cases.
	As a response to these issues, we introduce a regularization method that
	exploits prior knowledge about the spatiotemporal nature of the \ac{eeg} signal to
	improve the accuracy for low data availability settings and speed up the
	classifier training time, thereby considerably reducing memory usage.
	Similar structured regularization approaches have been applied to other linear
	\ac{erp} classifiers~\cite{GonzalezNavarro2017, Vliet2020} and have shown
	significant increases in performance.
	Additionally, we show that regularization results in an interpretable
	classification model, which can aid in analyzing and developing spatiotemporal beamformer-based classifiers.

	\section{Materials \& methods}
	\subsection{Notation}
	We represent matrices with bold capital letters, vectors with bold
	lowercase letters, fixed scalars with uppercase cursive letters and variable
  scalars with cursive lowercase letters.
	The epoched \ac{eeg} data with $N$ epochs, $C$ channels, and $S$ samples are
  represented in epoch format as $\{\mat{X}_i\in\mathbb{R}^{C\times S}\}^N_{i=1}$
  or flattened vector format by concatenating all channels for each epoch.
	Flattening results in $\{\mat{x}_i\in\mathbb{R}^{CS}\}^N_{i=1}$ such that
  $\mat{x}_i = \text{vec}(\mat{X}_i)$.
	The real covariance matrix of the epochs in vector format is
  denoted by $\mat{C}$, estimators thereof as $\hat{\mat{C}}$.

	\subsection{Spatiotemporal beamforming}
  \ac{lcmv}-beamforming was initially introduced to \ac{eeg} analysis as a filter for
	source localization~\cite{VanVeen1997} to enhance the signal-to-noise ratio
	(SNR).
	\textcite{VanVliet2015} first applied the spatiotemporal
	\ac{lcmv}-beamformer as a method for the analysis of \acp{erp}.
	The extension to the combined spatiotemporal domain~\cite{VanVliet2015} and the
	data-driven approaches proposed by Treder et al.~\cite{Treder2016} and
	Wittevrongel et al.~\cite{Wittevrongel2016} allow for its application to classification problems.

	For the following analyses, we assume that all \ac{eeg} channels are normalized with zero mean and unit variance without loss of generality.
	Solving \cref{eq:minimum_variance} under the linear constraint given by
	\cref{eq:linear_constraint} returns the filter weights $w$ defining the spatiotemporal \ac{lcmv}-beamformer.
	\begin{equation}
    \underset{\mat{w}}{\arg\min}\mat{w}^\intercal \mat{C}
		\mat{w}^\intercal
		\label{eq:minimum_variance}
	\end{equation}
	\begin{equation}
		\mat{a}^\intercal\mat{w} = 1
		\label{eq:linear_constraint}
	\end{equation}

	These weights minimize the variance of the output of the filter while enhancing
	the signal characterized by the constraint.
  $\mat{a} = \text{vec}(\mat{A})$ is the data-driven activation pattern, a template
	of the signal of interest maximizing the difference between two classes of
	epochs, determined as follows:

@@

	The method of Lagrange multipliers then gives the closed-form solution to the minimization problem posed by
	\cref{eq:minimum_variance} and \cref{eq:linear_constraint} as:

	\begin{equation}
		\mat{w} =
    \frac{\mat{C}^{-1}\mat{a}^\intercal}
		{\mat{a}C^{-1}\mat{a}^\intercal}
		\label{eq:closed_form}
	\end{equation}

@@

	resulting in a scalar output per epoch.
	The linear constraint in \cref{eq:linear_constraint} ensures that the
	beamformer maps epochs containing a target response to a score close to one
	and, conversely, epochs not containing a target response to a score close to
	zero.

	\subsection{Covariance matrix regularization}
	While the spatiotemporal beamformer, in theory, achieves optimal separation
	between target and non-target classes, in analogy to linear discriminant
  analysis~\cite{Treder201}, it does not always perform well on unseen data.
	The main challenge is to find a good estimator for the inverse covariance
  matrix $\mat{C}^{-1}$ since the real underlying covariance matrix generating the data is, in principle, unknown.

@@

	\begin{equation}
    \hat{\mat{C}}_\text{emp} =
		\frac{1}{N-1}\sum^{N}_{i=1}\mat{x}_i\mat{x}_i^\intercal
		\label{eq:empirical_covariance}
	\end{equation}

	\begin{equation}
    \widehat{\mat{C}^{-1}}_\text{emp} = \hat{\mat{C}}_\text{emp}^+
		\label{eq:empirical_inverse_covariance}
	\end{equation}

	The Moore-Penrose pseudoinverse $^+$, ensures a solution exists when
  $\hat{\mat{C}}_\text{emp}$ is singular.
	\cref{fig:kronlda-covs}a and \cref{fig:kronlda-covs}b respectively show examples of the
	empirical estimators of the covariance and the inverse covariance matrices.
	The empirical estimator suffers from performance and
	stability issues if the number of epochs $n$ used or estimation is not much larger than the number of features $cs$~\cite{Stein1956,Khatri1987}.

@@

	\begin{equation}
    \hat{\mat{C}}_\alpha =
    (1-\alpha) \hat{\mat{C}}_\text{emp}
    + \alpha\frac{\text{Tr}(\hat{\mat{C}}_\text{emp})}{CS}\mathbb{I}
		\label{eq:shrinkage}
	\end{equation}

	\begin{equation}
    \widehat{\mat{C}^{-1}}_\alpha =
    \hat{\mat{C}}^+_\alpha
		\label{eq:shrinkage_inverse}
	\end{equation}

	with $0<\alpha<1$.
	Analogous to $L_2$ regularization of the beamforming problem,
	shrinkage reduces the ratio between the smallest and largest eigenvalues
	of the covariance matrix by strengthening the diagonal.
	\cref{fig:kronlda-covs}c and \cref{fig:kronlda-covs}d respectively show examples of the
	shrunk estimator of the covariance and the inverse covariance matrices.

	Earlier work~\cite{Libert2021} applied shrinkage regularization to \ac{erp}
	decoding with the spatiotemporal beamformer and showed competitive performance
	compared to other state-of-the-art decoding techniques like stepwise LDA or SVM.
	The abovementioned work chooses the shrinkage coefficient $\alpha$ as a fixed hyperparameter.
	However, its optimal value depends on the number of training epochs, the
	covariance matrix's dimensionality, and the independence and variance of the
	data, which can vary across evaluation settings and per session.
	The optimal value for $\alpha$ can be found with a line search using cross-validation, but this can be a costly procedure.
	Methods exist to estimate an optimal shrinkage value from the data directly.
	Most notable among these are the Ledoit-Wolf procedure~\cite{Ledoit2004},
	Rao-Blackwell Ledoit-Wolf~\cite{Chen2010}, and Oracle Approximating Shrinkage~\cite{Chen2010}.
	A more recent estimation method~\cite{Tong2018} emulates a leave-one-out
	cross-validation (\ac{loocv}) scheme expressed by the data-driven closed-form
	estimate:

	\begin{equation}
		\alpha =
		1-\frac{
      \frac{N}{N-1}\text{Tr}(\hat{\mat{C}}_\text{emp}^2)
      - \frac{2}{CS}\left[\text{Tr}(\hat{\mat{C}}_\text{emp})\right]^2
      + \frac{1}{CS}\text{Tr}(\hat{\mat{C}}_\text{emp}^2)
			- \frac{1}{N(N-1)}\sum_{i=1}^N||\mat{x}_i||_2^4
		}
		{
      \frac{N^2 -2N}{(N-1)^2}\text{Tr}(\hat{\mat{C}}_\text{emp}^2)
      - \frac{2}{CS}\left[\text{Tr}(\hat{\mat{C}}_\text{emp})\right]^2
      + \frac{1}{CS}\text{Tr}(\hat{\mat{C}}_\text{emp}^2)
			+ \frac{1}{N(N-1)^2}\sum_{i=1}^N||\mat{x}_i||_2^4
		}
		\label{eq:loocv}
	\end{equation}

	We opt for the \ac{loocv} shrinkage estimator because it avoids some of the
	assumptions made by~\cite{Ledoit2004} and~\cite{Chen2010} and
	because it generalizes to structured covariance estimation as described in
	\cref{seq:structured_estimation}.

	\subsubsection{Spatiotemporal beamforming with Kronecker-Toeplitz structured covariance}
	\label{seq:structured_estimation}
	Exploiting prior knowledge about the spatiotemporal structure of the \ac{eeg} signal leads to a more regularized estimator of the covariance.
	When viewing the example of empirical spatiotemporal \ac{eeg} covariance in
	\cref{fig:kronlda-covs}a, it becomes clear that this matrix consists of a block pattern of repeated, similar matrices.
	Due to the multi-channel nature of the signal, we assume that the covariance of spatiotemporal \ac{eeg} epochs is a Kronecker
	product of two smaller
	matrices~\cite{Munck1992,DeMunck1999,Huizenga2002}, as expressed
	by:

	\begin{equation}
    \hat{\mat{C}}_\text{struct} = \hat{\mat{S}} \otimes \hat{\mat{T}}
		\label{eq:kronecker}
	\end{equation}

	with $\otimes$ the Kronecker product operator.
  $\hat{\mat{S}} \in \mathbb{R}^{C\times C}$ and $\hat{\mat{T}} \in \mathbb{R}^{S\times S}$ respectively correspond to estimators of the spatial and temporal covariance of the data.
	Furthermore, because the temporal covariance of the \ac{eeg}-signal is
	stationary (i.e., it is only dependent on interval length between covarying
	time samples)~\cite{Bijma2003}, it is assumed to have a Toeplitz-matrix structure:

	\begin{equation}
    \hat{\mat{T}}_{i,j} = \hat{\mat{T}}_{i+1,j+1}
		\label{eq:toeplitz}
	\end{equation}

	\Cref{prop:inverse_kronecker} then leads to
	\cref{eq:cov_inverse_kronecker} to estimate the inverse
	covariance.
	\begin{property}
    $(\mat{U} \otimes \mat{V})^+ = \mat{U}^+ \otimes \mat{V}^+$ for any non-singular
    matrices $\mat{U}$ and $\mat{V}$~\cite{Langville2004}.
		\label{prop:inverse_kronecker}
	\end{property}

	\begin{equation}
    \widehat{\mat{C}^{-1}}_\text{struct} = \hat{\mat{S}}^+ \otimes \hat{\mat{T}}^+
		\label{eq:cov_inverse_kronecker}
	\end{equation}

	Finally, based on \cref{prop:kronecker_multiplication},
	\cref{eq:closed_form} can be reformulated more efficiently as
	\cref{eq:closed_form_kronecker}.
	\begin{property}
    $(\mat{U}\otimes \mat{V})\cdot\text{vec}(\mat{W}) =
    \text{vec}(\mat{V}\mat{W}\mat{U}^\intercal)$
    for any matrices $\mat{U}\in\mathbb{R}^{P\times P}$,
    $\mat{V}\in\mathbb{R}^{Q\times Q}$ and $\mat{W}\in\mathbb{R}^{P\times Q}$~\cite{Loan2000}.
		\label{prop:kronecker_multiplication}
	\end{property}

	\begin{equation}
		\hat{\mat{w}}_\text{struct} =
    \frac{\hat{\mat{S}}^+\mat{A}^\intercal\hat{\mat{T}}^+}
    {\mat{a}\cdot\text{vec}(\hat{\mat{S}}^+A^\intercal\hat{\mat{T}}^+)}
		\label{eq:closed_form_kronecker}
	\end{equation}

	Using \cref{eq:closed_form_kronecker} removes the need to calculate the
  full, high dimensional Kronecker product $\hat{\mat{S}}^+\otimes
    \hat{\mat{T}}^+$.
	\Cref{fig:kronlda-covs}e and \cref{fig:covs}f respectively show examples of the
	structured covariance and inverse covariance estimators,
	consisting of a spatial Kronecker factor (\cref{fig:kronlda-covs}g and
	\cref{fig:kronlda-covs}h) and a temporal component (\cref{fig:kronlda-covs}i and
	\cref{fig:kronlda-covs}j).

	\begin{figure}
		\includegraphics[width=\linewidth]{figures/stbf_struct/covs.eps}
    \caption[Estimated covariance and inverse covariance.]{Different estimators of the covariance and inverse covariance
			of 100 epochs of data from \textit{Subject 01} for channels
			\textit{Fz}, \textit{Cz}, \textit{Pz}, and \textit{Oz} and time samples between 0.1s and 0.6s.
			Regularized estimators of the inverse covariance exhibit less extreme values and have a sparser structure.
			(\textbf{a,b}) Empirical covariance and inverse covariance matrices.
			(\textbf{c,d}) Shrunk covariance and inverse covariance matrices with $\alpha=0.14$ as
			determined by the closed-form \ac{loocv} method. (\textbf{e,f}) Kronecker-Toeplitz
			structured covariance and inverse covariance matrices.
			(\textbf{g,h}) Spatial Kronecker factor of the Kronecker-Toeplitz structured shrunk estimator and its inverse.
			(\textbf{i,j}) Temporal Kronecker factor of the Kronecker-Toeplitz structured shrunk estimator and its inverse.}
		\label{fig:kronlda-covs}
	\end{figure}

	The Kronecker approach has shown significant performance yields in different linear spatiotemporal \ac{eeg} and MEG
	applications~\cite{DeMunck2002,Huizenga2002,Beltrachini2013,GonzalezNavarro2016,GonzalezNavarro2017}.
	\textcite{Vliet2020} have applied a Kronecker-structured covariance estimator to \ac{erp} classification with linear models in a post-hoc fashion.
	Our work goes further by embedding the Kronecker structure in the
	spatiotemporal beamformer training process, using a data-adaptive shrinkage
	method, and regularizing the covariance further by imposing a Toeplitz
	structure on the temporal covariance.

@@

	\begin{subequations}
		\begin{equation}
      \tilde{\mat{S}}_{k+1} =
			\frac{1}{N}
      \sum^N_{i=1}\mat{X}_i^\intercal\hat{\mat{T}}_k^+\mat{X}_i
			\label{eq:fpi_spatial}
		\end{equation}
		\begin{equation}
      \tilde{\mat{T}}_{k+1} =
			\frac{1}{N}
      \sum^N_{i=1}\mat{X}_i\hat{\mat{S}}_k^+\mat{X}_i^\intercal
			\label{eq:fpi_temporal}
		\end{equation}
	\end{subequations}

	\begin{subequations}
		\begin{equation}
      \tilde{\mat{S}}_{k+1}^{(\beta)} =
      (1-\beta_{k+1})\tilde{\mat{S}}_{k+1}
      +\beta_{k+1}\frac{\text{Tr}(\tilde{\mat{S}}_{k+1})}{C}\mathbb{I}
			\label{eq:fpi_spatial_shrunk}
		\end{equation}
		\begin{equation}
      \tilde{\mat{T}}_{k+1}^{(\gamma)} =
      (1-\gamma_{k+1})\tilde{\mat{T}}_{k+1}
      +\gamma_{k+1}\frac{\text{Tr}(\tilde{\mat{T}}_{k+1})}{S}\mathbb{I}
			\label{eq:fpi_temporal_shrunk}
		\end{equation}
	\end{subequations}

	\begin{subequations}
		\begin{equation}
      \hat{\mat{S}}_{k+1} =
      \frac{C}{\text{Tr}\left[\tilde{\mat{S}}_{k+1}^{(\beta)}\right]}
      \tilde{\mat{S}}_{k+1}^{(\beta)}
			\label{eq:fpi_spatial_norm}
		\end{equation}
		\begin{equation}
      \hat{\mat{T}}_{k+1} =
      \frac{S}{\text{Tr}\left[\tilde{\mat{T}}_{k+1}^{(\gamma)}\right]}
      \tilde{\mat{T}}_{k+1}^{(\gamma)}
			\label{eq:fpi_temporal_norm}
		\end{equation}
	\end{subequations}
